---
title: "Toy MIRAGE Examples"
output: workflowr::wflow_html
---

## Toy Input Format

MIRAGE expects one row per rare variant. The table below uses fake toy data only.

```{r}
toy_variants <- data.frame(
  ID = paste0("var", 1:10),
  Gene = c("GENE1", "GENE1", "GENE1", "GENE2", "GENE2",
           "GENE3", "GENE3", "GENE4", "GENE4", "GENE4"),
  No.case = c(3, 1, 0, 4, 2, 0, 1, 5, 1, 0),
  No.contr = c(0, 1, 2, 1, 2, 1, 1, 0, 2, 1),
  category = c(1, 2, 3, 1, 2, 3, 2, 1, 2, 3)
)

toy_variants
```

Example category labels:

```{r}
category_key <- data.frame(
  category = 1:3,
  label = c("LoF-like", "damaging-missense-like", "other-rare")
)

category_key
```

## Example Workflow

The code below shows how a MIRAGE run would be structured. It is not evaluated here because this website should remain lightweight and independent of package installation state.

```{r, eval=FALSE}
library(mirage)

result <- mirage(
  toy_variants,
  n1 = 500,
  n2 = 500,
  gamma = c(6, 3, 2),
  sigma = 2,
  eta.init = 0.1,
  delta.init = 0.05,
  verbose = FALSE
)

head(result$BF.PP.gene)
result$eta.est
```

## Minimal Sanity Checks

Before running MIRAGE on real summary statistics, use checks like these on the input table.

```{r}
stopifnot(all(c("ID", "Gene", "No.case", "No.contr", "category") %in% names(toy_variants)))
stopifnot(all(toy_variants$No.case >= 0))
stopifnot(all(toy_variants$No.contr >= 0))
stopifnot(!anyDuplicated(toy_variants$ID))
```
