---
title: "Toy MIRAGE Examples"
output: workflowr::wflow_html
---

## Data Policy

This page uses fake toy data only. It does not use real WES, ASD, dbGaP, SPARK, protected, private, or large data.

## Expected Input Columns

For gene-level MIRAGE analysis, a toy input table should contain one row per rare variant:

- `ID`: variant identifier.
- `Gene`: gene name or gene identifier.
- `No.case`: allele count in cases.
- `No.contr`: allele count in controls.
- `category`: annotation group index.

```{r}
toy_variants <- data.frame(
  ID = paste0("toy_var_", seq_len(12)),
  Gene = c("GENE1", "GENE1", "GENE1", "GENE2", "GENE2", "GENE2",
           "GENE3", "GENE3", "GENE4", "GENE4", "GENE4", "GENE4"),
  No.case = c(3, 1, 0, 4, 2, 0, 0, 1, 5, 1, 0, 2),
  No.contr = c(0, 1, 2, 1, 2, 1, 1, 1, 0, 2, 1, 2),
  category = c(1, 2, 3, 1, 2, 3, 3, 2, 1, 2, 3, 2)
)

toy_variants
```

```{r}
category_key <- data.frame(
  category = 1:3,
  toy_label = c("LoF-like", "damaging-missense-like", "other-rare")
)

category_key
```

## Basic Input Checks

```{r}
required_columns <- c("ID", "Gene", "No.case", "No.contr", "category")

stopifnot(all(required_columns %in% names(toy_variants)))
stopifnot(all(toy_variants$No.case >= 0))
stopifnot(all(toy_variants$No.contr >= 0))
stopifnot(!anyDuplicated(toy_variants$ID))
stopifnot(all(toy_variants$category %in% category_key$category))
```

## Optional MIRAGE Run

If the `mirage` package is available in the local R library, this page runs a very small toy example. If it is not available, the page prints the command that would be used.

```{r}
mirage_available <- requireNamespace("mirage", quietly = TRUE)
mirage_available
```

```{r}
if (mirage_available) {
  toy_result <- mirage::mirage(
    toy_variants,
    n1 = 500,
    n2 = 500,
    gamma = c(6, 3, 2),
    sigma = 2,
    eta.init = 0.1,
    delta.init = 0.05,
    max.iter = 100,
    verbose = FALSE
  )

  head(toy_result$BF.PP.gene)
  toy_result$eta.est
} else {
  message("The mirage package is not installed locally. See the non-evaluated example below.")
}
```

```{r, eval=FALSE}
library(mirage)

toy_result <- mirage(
  toy_variants,
  n1 = 500,
  n2 = 500,
  gamma = c(6, 3, 2),
  sigma = 2,
  eta.init = 0.1,
  delta.init = 0.05,
  verbose = FALSE
)

head(toy_result$BF.PP.gene)
toy_result$eta.est
toy_result$delta.est
```

## Expected Output Objects

The gene-level `mirage()` result is expected to include:

- `BF.PP.gene`: gene-level Bayes factors and posterior probabilities.
- `BF.all`: variant-level information and Bayes factors grouped by gene.
- `eta.est`: estimated category-specific risk-variant proportions.
- `delta.est`: estimated or fixed gene-level prior probability, depending on settings.

## Interpretation Notes

- A larger gene BF indicates stronger model-based evidence for the risk-gene model over the non-risk-gene model.
- A larger posterior probability indicates higher fitted probability that the gene is a risk gene under the model.
- Larger `eta` values suggest that variants in that annotation category are estimated to be more often risk variants.
- Bayesian FDR can be summarized by averaging \(1 - \operatorname{PP}_i\) across a selected set of genes.
- These quantities are model-based and should be interpreted alongside sensitivity analyses and data-quality checks.
