---
title: "Simulation"
output: workflowr::wflow_html
---

## Toy Simulation Only

This page runs a small synthetic software smoke check using the current implementation. It does not use real ASD, schizophrenia, WES, dbGaP, SPARK, protected, private, or large data.

The goal is to verify that the code runs end-to-end, converges, and separates a strong synthetic pleiotropic signal from non-pleiotropic genes. This is not a formal parameter-recovery, power, or FDR-calibration study.

```{r}
source("R/two_trait_mirage.R")
source("R/simulate_two_trait_mirage.R")

set.seed(20260612)
sim <- simulate_two_trait_mirage(
  n_genes = 120,
  variants_per_gene = 8,
  pi = c("00" = 0.70, "10" = 0.10, "01" = 0.08, "11" = 0.12),
  eta_trait1 = c("1" = 0.45, "2" = 0.10),
  eta_trait2 = c("1" = 0.45, "2" = 0.10),
  signal_mean_trait1 = 3.5,
  signal_mean_trait2 = 3.5,
  null_mean = -1.5,
  seed = 20260612
)

fit <- mirage_two_trait(
  sim$trait1,
  sim$trait2,
  n1_trait1 = 500,
  n0_trait1 = 500,
  n1_trait2 = 500,
  n0_trait2 = 500,
  log_bf_col_trait1 = "log_bf",
  log_bf_col_trait2 = "log_bf",
  eta_init_trait1 = c("1" = 0.20, "2" = 0.05),
  eta_init_trait2 = c("1" = 0.20, "2" = 0.05),
  pi_init = c("00" = 0.75, "10" = 0.08, "01" = 0.07, "11" = 0.10),
  max_iter = 100,
  verbose = FALSE
)
```

## Fitted Parameters

```{r}
fit$parameters$pi
fit$parameters$eta_trait1
fit$parameters$eta_trait2
```

## True Versus Estimated Parameters

```{r}
pi_comparison <- data.frame(
  parameter = names(sim$parameters$pi),
  truth = as.numeric(sim$parameters$pi),
  estimate = as.numeric(fit$parameters$pi[names(sim$parameters$pi)])
)

eta_comparison <- rbind(
  data.frame(
    trait = "trait1",
    category = names(sim$parameters$eta_trait1),
    truth = as.numeric(sim$parameters$eta_trait1),
    estimate = as.numeric(fit$parameters$eta_trait1[names(sim$parameters$eta_trait1)])
  ),
  data.frame(
    trait = "trait2",
    category = names(sim$parameters$eta_trait2),
    truth = as.numeric(sim$parameters$eta_trait2),
    estimate = as.numeric(fit$parameters$eta_trait2[names(sim$parameters$eta_trait2)])
  )
)

pi_comparison
eta_comparison
```

The estimates in this single toy run are expected to differ from the generating values. A replicated simulation grid is still needed before making claims about parameter recovery or Bayesian FDR calibration.

## Diagnostics

```{r}
data.frame(
  converged = fit$diagnostics$converged,
  iterations = fit$diagnostics$n_iter,
  loglik_non_decreasing = fit$diagnostics$loglik_non_decreasing,
  final_loglik = tail(fit$traces$loglik, 1)
)
```

## Pleiotropy Posterior Check

```{r}
truth <- sim$truth[match(fit$posterior$Gene, sim$truth$Gene), ]
data.frame(
  group = c("true_pleiotropic", "not_true_pleiotropic"),
  mean_PP_pleiotropy = c(
    mean(fit$posterior$PP_pleiotropy[truth$state == "11"]),
    mean(fit$posterior$PP_pleiotropy[truth$state != "11"])
  )
)
```

## Top Pleiotropy Candidates In Toy Data

```{r}
head(fit$bfdr$pleiotropy, 10)
```

These numbers are generated from synthetic data and should be interpreted only as a software smoke check, not as evidence of final calibration.
