---
title: "Simulation"
output: workflowr::wflow_html
---

## Toy Simulation Only

This page runs a small synthetic software smoke check using the current implementation. It does not use real ASD, schizophrenia, WES, dbGaP, SPARK, protected, private, or large data.

The goal is to verify that the code runs end-to-end, converges, and separates a strong synthetic pleiotropic signal from non-pleiotropic genes. This is not a formal parameter-recovery, power, or FDR-calibration study.

```{r}
source("R/two_trait_mirage.R")
source("R/simulate_two_trait_mirage.R")

set.seed(20260612)
sim <- simulate_two_trait_mirage(
  n_genes = 120,
  variants_per_gene = 8,
  pi = c("00" = 0.70, "10" = 0.10, "01" = 0.08, "11" = 0.12),
  eta_trait1 = c("1" = 0.45, "2" = 0.10),
  eta_trait2 = c("1" = 0.45, "2" = 0.10),
  signal_mean_trait1 = 3.5,
  signal_mean_trait2 = 3.5,
  null_mean = -1.5,
  seed = 20260612
)

fit <- mirage_two_trait(
  sim$trait1,
  sim$trait2,
  n1_trait1 = 500,
  n0_trait1 = 500,
  n1_trait2 = 500,
  n0_trait2 = 500,
  log_bf_col_trait1 = "log_bf",
  log_bf_col_trait2 = "log_bf",
  eta_init_trait1 = c("1" = 0.20, "2" = 0.05),
  eta_init_trait2 = c("1" = 0.20, "2" = 0.05),
  pi_init = c("00" = 0.75, "10" = 0.08, "01" = 0.07, "11" = 0.10),
  max_iter = 100,
  verbose = FALSE
)
```

## Fitted Parameters

```{r}
fit$parameters$pi
fit$parameters$eta_trait1
fit$parameters$eta_trait2
```

## True Versus Estimated Parameters

```{r}
pi_comparison <- data.frame(
  parameter = names(sim$parameters$pi),
  truth = as.numeric(sim$parameters$pi),
  estimate = as.numeric(fit$parameters$pi[names(sim$parameters$pi)])
)

eta_comparison <- rbind(
  data.frame(
    trait = "trait1",
    category = names(sim$parameters$eta_trait1),
    truth = as.numeric(sim$parameters$eta_trait1),
    estimate = as.numeric(fit$parameters$eta_trait1[names(sim$parameters$eta_trait1)])
  ),
  data.frame(
    trait = "trait2",
    category = names(sim$parameters$eta_trait2),
    truth = as.numeric(sim$parameters$eta_trait2),
    estimate = as.numeric(fit$parameters$eta_trait2[names(sim$parameters$eta_trait2)])
  )
)

pi_comparison
eta_comparison
```

The estimates in this single toy run are expected to differ from the generating values. A replicated simulation grid is still needed before making claims about parameter recovery or Bayesian FDR calibration.

## Diagnostics

```{r}
data.frame(
  converged = fit$diagnostics$converged,
  iterations = fit$diagnostics$n_iter,
  loglik_non_decreasing = fit$diagnostics$loglik_non_decreasing,
  final_loglik = tail(fit$traces$loglik, 1)
)
```

## Pleiotropy Posterior Check

```{r}
truth <- sim$truth[match(fit$posterior$Gene, sim$truth$Gene), ]
data.frame(
  group = c("true_pleiotropic", "not_true_pleiotropic"),
  mean_PP_pleiotropy = c(
    mean(fit$posterior$PP_pleiotropy[truth$state == "11"]),
    mean(fit$posterior$PP_pleiotropy[truth$state != "11"])
  )
)
```

## Top Pleiotropy Candidates In Toy Data

```{r}
head(fit$bfdr$pleiotropy, 10)
```

These numbers are generated from synthetic data and should be interpreted only as a software smoke check, not as evidence of final calibration.

## mTADA-Style Benchmark

Nguyen et al. (2020) evaluated mTADA by assigning each gene to one of four
two-trait states and simulating de novo mutation counts from Poisson models.
Their main design used 19,538 genes, two mutation categories, single-trait risk
proportions of 0.05 and 0.03, a grid of pleiotropic proportions from zero to
0.03, multiple sample-size and mean-relative-risk settings, and 100 replicates
per setting.

The repository now includes `R/mtada_style_simulation.R`, which reproduces this
data-generating structure using synthetic mutation rates. For each
gene-category-trait combination it:

1. samples a four-state gene label;
2. samples a relative risk from the trait- and category-specific Gamma prior
   for active traits;
3. samples a de novo count from the corresponding Poisson distribution; and
4. computes the analytic TADA de novo log Bayes factor.

The de novo log Bayes factors are passed to the existing two-trait MIRAGE
four-state engine. `eta` is fixed to one in this benchmark, so each simulated
gene-category Bayes factor enters directly. This isolates the joint gene-state
inference and avoids claiming that mTADA gene-level counts are variant-level
case-control observations.

```{r mtada-style-example}
source("R/mtada_style_simulation.R")

mtada_sim <- simulate_mtada_style(
  n_genes = 500,
  pi_trait1 = 0.05,
  pi_trait2 = 0.03,
  pi_pleiotropic = 0.015,
  n_trios_trait1 = 5122,
  n_trios_trait2 = 1077,
  seed = 20260619
)

mtada_result <- fit_mtada_style_mirage(mtada_sim)
mtada_result$metrics$auc
mtada_result$metrics$pp_threshold
mtada_result$metrics$bfdr
mtada_result$metrics$pi
```

The benchmark reports joint and separate-trait AUC, discoveries and empirical
FDP at posterior probability 0.8, pleiotropic classification, and recovery of
the four mixing proportions. A configurable replicated grid is available
through `run_mtada_style_grid()`. These outputs follow the main comparison
strategy in the mTADA paper, but remain a synthetic method-development
benchmark rather than a reproduction of the paper's real mutation-rate data.

## Paper-Profile Pilot Result

The script `code/run_mtada_style_benchmark.R` was run once with 19,538 genes,
two categories, 5,122 trait-1 trios, 1,077 trait-2 trios, and
\((\pi_1^S,\pi_2^S,\pi_3)=(0.05,0.03,0.015)\). The seed was 20260619.

```{r mtada-paper-profile-result}
data.frame(
  target = c("trait1", "trait1", "trait2", "trait2", "pleiotropy"),
  method = c("joint", "separate", "joint", "separate", "joint"),
  AUC = c(0.9968, 0.9968, 0.8323, 0.7024, 0.9838)
)

data.frame(
  state = c("00", "10", "01", "11"),
  truth = c(0.935, 0.035, 0.015, 0.015),
  estimate = c(0.9336, 0.0380, 0.0144, 0.0140)
)
```

At posterior probability 0.8, the joint analysis found 47 trait-2 genes
compared with 26 from the separate analysis; their observed FDPs were 0.021
and 0, respectively. The joint trait-2 power was 0.082 versus 0.046 for the
separate analysis. Trait 1 was already highly informative and changed little.
The pleiotropy analysis selected 32 genes, with power 0.109 and observed FDP
0.031.

This is one favorable replicate, not a power or calibration conclusion. The
next formal analysis should run the full grid over pleiotropic proportion,
sample size, mean relative risk, and repeated seeds, then summarize the
distribution of AUC, power, observed FDP, and parameter error.

## Full Paper Grid

The full grid has now been encoded in `mtada_paper_grid()`:

| Dimension | Values |
|---|---|
| Shared-risk proportion | \(\pi_{11}=0, 0.02\) |
| Trait 1 mRR | \((105,29)\) or \((24,6)\) |
| Trait 2 mRR | \((12,2)\) or \((87,23)\) |
| Trio counts | All pairs from 1,000, 2,000, and 5,000 |
| Repeated seeds | 100 per setting |

This produces \(2\times4\times9\times100=7{,}200\) independently generated
datasets. The generator uses the 19,358 gene-category mutation rates distributed
with the public mTADA repository when that external reference directory is
provided. The mutation-rate file remains untracked and is not copied into this
repository.

Each dataset is evaluated with:

1. the joint four-state MIRAGE likelihood;
2. separate single-trait posterior analyses; and
3. the original mTADA `DN2traits` Stan model, using its public posterior
   calculation code.

Recorded metrics are AUC, power and empirical FDP at PP at least 0.8, observed
FDP versus estimated Bayesian FDR at targets 0.01, 0.05, 0.10, and 0.20,
\(\pi_{11}\) recovery, and elapsed runtime. The runner writes one checkpoint per
task and supports non-overlapping task ranges for parallel or resumed runs.

The original mTADA implementation was verified under R 4.4.2 with rstan 2.32.7
and locfit 1.5-9.12. Compilation took about 85 seconds. A full 19,358-gene,
5,000-iteration VB fit took about 209 seconds in a reliability check, compared
with about 6 seconds for the joint MIRAGE fit in that same run. Consequently,
the complete original-mTADA grid is a multi-day computation even with several
workers. No full-grid performance claim should be made until all 7,200
checkpoints have completed and been aggregated.

### Reference

Nguyen TH, Dobbyn A, Brown RC, et al. mTADA is a framework for identifying risk
genes from de novo mutations in multiple traits. *Nature Communications*.
2020;11:2929. doi:10.1038/s41467-020-16487-z.
