---
title: "Simulation"
output: workflowr::wflow_html
---

## Study Objective

This simulation evaluates the two-trait MIRAGE extension using the same
gene-level data-generating mechanism used in the mTADA paper. The comparison is
performed on identical simulated datasets so that differences reflect the
analysis methods rather than different data generation.

This page distinguishes two types of information:

- **Taken from the public mTADA repository:** gene names and mutation rates for
  19,358 genes in two annotation categories.
- **Simulated for this study:** four-state gene status, relative risks, and de
  novo mutation counts for the two traits.

The observed disease mutation counts distributed with mTADA are not used as
simulation outcomes. No ASD, schizophrenia, WES, dbGaP, SPARK, protected, or
private individual-level data are included in this repository.

## Data Source

The mutation-rate profile is read locally from the public mTADA file:

```text
mTADA/data/FullDataSet_DenovoMutations_for_mTADA.txt
```

Only these columns are used:

| Field | Use |
|---|---|
| `Gene` | Common gene identifier |
| `mut_damaging` | Mutation rate for the damaging missense category |
| `mut_lof` | Mutation rate for the loss-of-function category |

The mTADA repository is treated as an external reference and is not copied into
this GitHub repository.

## Data-Generating Mechanism

For every gene \(i\), a gene-level state is sampled:

\[
S_i \in \{00,10,01,11\}.
\]

The state probabilities are derived from the single-trait risk proportions
\(\pi_1^S\), \(\pi_2^S\), and the shared-risk proportion \(\pi_{11}\):

\[
\begin{aligned}
\pi_{00} &= 1-\pi_1^S-\pi_2^S+\pi_{11},\\
\pi_{10} &= \pi_1^S-\pi_{11},\\
\pi_{01} &= \pi_2^S-\pi_{11},\\
\pi_{11} &= \pi_{11}.
\end{aligned}
\]

As in the mTADA simulation, \(\pi_1^S=0.05\) and \(\pi_2^S=0.03\).

For an active trait \(t\) and annotation category \(c\), the gene-specific
relative risk is generated from:

\[
\gamma_{itc}\sim
\operatorname{Gamma}(\bar{\gamma}_{tc}\beta_{tc},\beta_{tc}).
\]

For an inactive trait, \(\gamma_{itc}=1\). Given mutation rate \(\mu_{ic}\),
trio count \(N_t\), and relative risk \(\gamma_{itc}\), the de novo count is:

\[
X_{itc}\sim
\operatorname{Poisson}(2N_t\mu_{ic}\gamma_{itc}).
\]

This is the mTADA de novo Poisson/Gamma mechanism. State \(11\) is assigned at
the gene level and does not require the same mutation category to drive both
traits.

## Analysis Workflow

For each simulated gene-category count, the analytic TADA de novo Bayes factor
is calculated:

\[
\mathrm{BF}_{itc}
=
\frac{
\operatorname{NB}\left(
X_{itc};
\bar{\gamma}_{tc}\beta_{tc},
\frac{\beta_{tc}}{\beta_{tc}+2N_t\mu_{ic}}
\right)
}{
\operatorname{Poisson}(X_{itc};2N_t\mu_{ic})
}.
\]

Category Bayes factors are combined within each gene on the log scale. The same
simulated counts and mutation rates are then analyzed by:

1. **Joint MIRAGE:** the proposed gene-level four-state model.
2. **Separate analyses:** independent single-trait posterior analyses.
3. **Original mTADA:** the public `DN2traits` Stan/VB implementation.

For this mTADA-based benchmark, category Bayes factors enter directly, which is
equivalent to fixing MIRAGE \(\eta=1\). This benchmark therefore tests the
joint four-state inference under the mTADA gene-category DGM. It does not
pretend that mTADA gene-level DNM counts are variant-level case-control data.

## Representative Run

One complete representative dataset was generated using the public mTADA
mutation rates and the following settings:

| Parameter | Value |
|---|---|
| Number of genes | 19,358 |
| Annotation categories | Damaging missense and LoF |
| Trait 1 risk proportion | 0.05 |
| Trait 2 risk proportion | 0.03 |
| Shared-risk proportion | 0.02 |
| Trait 1 trios | 2,000 |
| Trait 2 trios | 2,000 |
| Trait 1 mean RRs | 24 and 6 |
| Trait 2 mean RRs | 12 and 2 |
| Gamma dispersion parameter | 1 for each category |
| Simulation seed | 20200648 |
| Original mTADA VB iterations | 5,000 |

The true four-state proportions for this setting are:

\[
(\pi_{00},\pi_{10},\pi_{01},\pi_{11})
=(0.94,0.03,0.01,0.02).
\]

## Representative Results

### AUC

```{r representative-auc, echo=FALSE}
data.frame(
  method = c(
    "Joint MIRAGE", "Separate", "Original mTADA",
    "Joint MIRAGE", "Separate", "Original mTADA",
    "Joint MIRAGE", "Separate", "Original mTADA"
  ),
  target = rep(c("Trait 1", "Trait 2", "Pleiotropy"), each = 3),
  AUC = c(
    0.6687, 0.6587, 0.6685,
    0.6463, 0.5839, 0.6462,
    0.6886, 0.6887, 0.6887
  )
)
```

In this replicate, joint analysis improved Trait 2 AUC from 0.584 to 0.646.
Trait 1 AUC increased from 0.659 to 0.669. Joint MIRAGE and original mTADA
produced very similar rankings because both use the same four-state likelihood
structure in this benchmark.

### PP Threshold of 0.8

```{r representative-pp08, echo=FALSE}
data.frame(
  method = c(
    "Joint MIRAGE", "Separate", "Original mTADA",
    "Joint MIRAGE", "Separate", "Original mTADA",
    "Joint MIRAGE", "Separate", "Original mTADA"
  ),
  target = rep(c("Trait 1", "Trait 2", "Pleiotropy"), each = 3),
  discoveries = c(53, 44, 54, 9, 3, 13, 9, 0, 13),
  power = c(
    0.0514, 0.0444, 0.0524,
    0.0133, 0.0050, 0.0183,
    0.0193, 0, 0.0265
  ),
  empirical_FDP = c(
    0.0377, 0, 0.0370,
    0.1111, 0, 0.1538,
    0.1111, 0, 0.1538
  )
)
```

Joint MIRAGE found more Trait 1 and Trait 2 genes than the separate analyses.
For Trait 2, discoveries increased from 3 to 9. This gain came with one false
discovery in this replicate, so both power and empirical FDP must be evaluated
over repeated simulations.

### Mixing Proportion and Runtime

```{r representative-pi-runtime, echo=FALSE}
data.frame(
  method = c("Joint MIRAGE", "Original mTADA"),
  true_pi11 = c(0.02, 0.02),
  estimated_pi11 = c(0.0217, 0.0279),
  runtime_seconds = c(11.0, 108.9)
)
```

The separate analyses do not estimate a joint \(\pi_{11}\). The original mTADA
Stan VB fit completed, but its legacy `locfit` HPD routine raised
`HPD interval not closed` for this replicate. The reported mTADA estimate uses
a bounded density-mode fallback applied to the original `p12` VB draws. This
fallback is recorded in the task metadata rather than hidden.

Runtime excludes the one-time Stan compilation cost, which was approximately
85 seconds on the development computer.

### Bayesian FDR Check

At target Bayesian FDR 0.05:

```{r representative-bfdr, echo=FALSE}
data.frame(
  method = c(
    "Joint MIRAGE", "Separate", "Original mTADA",
    "Joint MIRAGE", "Separate", "Original mTADA"
  ),
  target = rep(c("Trait 1", "Trait 2"), each = 3),
  estimated_BFDR = c(
    0.0495, 0.0492, 0.0497,
    0.0358, 0.0439, 0.0410
  ),
  empirical_FDP = c(
    0.0217, 0, 0.0200,
    0.1250, 0, 0.1111
  )
)
```

This table is a diagnostic for one dataset, not an FDR-calibration result.
Calibration requires averaging empirical FDP across many replicates at each
simulation setting.

## Full mTADA Paper Grid

The implemented full grid follows the parameter combinations shown in the
mTADA paper:

| Dimension | Values |
|---|---|
| Shared-risk proportion | \(\pi_{11}=0, 0.02\) |
| Trait 1 mean RRs | \((105,29)\) or \((24,6)\) |
| Trait 2 mean RRs | \((12,2)\) or \((87,23)\) |
| Trio counts | All nine pairs formed from 1,000, 2,000, and 5,000 |
| Repeated seeds | 100 per setting |

The full design contains:

\[
2\times4\times9\times100=7{,}200
\]

simulated datasets. For every dataset the pipeline records:

- AUC for Trait 1, Trait 2, and pleiotropy;
- power and empirical FDP at PP at least 0.8;
- estimated BFDR and empirical FDP at targets 0.01, 0.05, 0.10, and 0.20;
- recovery of \(\pi_{11}\);
- runtime and fitting errors.

The runner uses one checkpoint per dataset and supports non-overlapping task
ranges for parallel execution. The full 7,200-dataset experiment has been
implemented but has not yet completed. Therefore, the representative result
above demonstrates that the data generation and all three analysis paths work;
it is not a final method comparison or calibration claim.

## Reproducible Commands

The full runner is `code/run_mtada_full_grid.R`. For example:

```{r full-grid-command, eval=FALSE}
Sys.setenv(
  MTADA_REFERENCE_DIR = "D:/path/to/mTADA",
  MIRAGE_TASK_START = "1",
  MIRAGE_TASK_END = "900",
  MIRAGE_RUN_ORIGINAL_MTADA = "true"
)
source("code/run_mtada_full_grid.R")
```

Completed checkpoints are summarized with:

```{r summarize-command, eval=FALSE}
source("code/summarize_mtada_full_grid.R")
```

## Internal Software Smoke Test

Before the mTADA-based benchmark was added, a small 120-gene synthetic test was
used only to verify EM convergence, posterior normalization, and end-to-end
execution. That internal test remains in the automated test suite, but it is
not the scientific simulation study reported on this page.

## Reference

Nguyen TH, Dobbyn A, Brown RC, et al. mTADA is a framework for identifying risk
genes from de novo mutations in multiple traits. *Nature Communications*.
2020;11:2929. doi:10.1038/s41467-020-16487-z.
