---
title: "Simulation"
output: workflowr::wflow_html
---

## Toy Simulation Only

This page runs a small synthetic smoke check using the current implementation. It does not use real ASD, schizophrenia, WES, dbGaP, SPARK, protected, private, or large data.

```{r}
source("R/two_trait_mirage.R")
source("R/simulate_two_trait_mirage.R")

set.seed(20260612)
sim <- simulate_two_trait_mirage(
  n_genes = 120,
  variants_per_gene = 8,
  pi = c("00" = 0.70, "10" = 0.10, "01" = 0.08, "11" = 0.12),
  eta_trait1 = c("1" = 0.45, "2" = 0.10),
  eta_trait2 = c("1" = 0.45, "2" = 0.10),
  signal_mean_trait1 = 3.5,
  signal_mean_trait2 = 3.5,
  null_mean = -1.5,
  seed = 20260612
)

fit <- mirage_two_trait(
  sim$trait1,
  sim$trait2,
  n1_trait1 = 500,
  n0_trait1 = 500,
  n1_trait2 = 500,
  n0_trait2 = 500,
  log_bf_col_trait1 = "log_bf",
  log_bf_col_trait2 = "log_bf",
  eta_init_trait1 = c("1" = 0.20, "2" = 0.05),
  eta_init_trait2 = c("1" = 0.20, "2" = 0.05),
  pi_init = c("00" = 0.75, "10" = 0.08, "01" = 0.07, "11" = 0.10),
  max_iter = 100,
  verbose = FALSE
)
```

## Fitted Parameters

```{r}
fit$parameters$pi
fit$parameters$eta_trait1
fit$parameters$eta_trait2
```

## Diagnostics

```{r}
data.frame(
  converged = fit$diagnostics$converged,
  iterations = fit$diagnostics$n_iter,
  loglik_non_decreasing = fit$diagnostics$loglik_non_decreasing,
  final_loglik = tail(fit$traces$loglik, 1)
)
```

## Pleiotropy Posterior Check

```{r}
truth <- sim$truth[match(fit$posterior$Gene, sim$truth$Gene), ]
data.frame(
  group = c("true_pleiotropic", "not_true_pleiotropic"),
  mean_PP_pleiotropy = c(
    mean(fit$posterior$PP_pleiotropy[truth$state == "11"]),
    mean(fit$posterior$PP_pleiotropy[truth$state != "11"])
  )
)
```

## Top Pleiotropy Candidates In Toy Data

```{r}
head(fit$bfdr$pleiotropy, 10)
```

These numbers are generated from synthetic data and should be interpreted only as a software smoke check.
