---
title: "Simulation"
output: workflowr::wflow_html
---

## Study Objective

This simulation evaluates the two-trait MIRAGE extension using the same
gene-level data-generating mechanism used in the mTADA paper. The comparison is
performed on identical simulated datasets so that differences reflect the
analysis methods rather than different data generation.

This page distinguishes two types of information:

- **Taken from the public mTADA repository:** gene names and mutation rates for
  19,358 genes in two annotation categories.
- **Simulated for this study:** four-state gene status, relative risks, and de
  novo mutation counts for the two traits.

The observed disease mutation counts distributed with mTADA are not used as
simulation outcomes. No ASD, schizophrenia, WES, dbGaP, SPARK, protected, or
private individual-level data are included in this repository.

## Data Source

The mutation-rate profile is read locally from the public mTADA file:

```text
mTADA/data/FullDataSet_DenovoMutations_for_mTADA.txt
```

Only these columns are used:

| Field | Use |
|---|---|
| `Gene` | Common gene identifier |
| `mut_damaging` | Mutation rate for the damaging missense category |
| `mut_lof` | Mutation rate for the loss-of-function category |

The mTADA repository is treated as an external reference and is not copied into
this GitHub repository.

## Data-Generating Mechanism

For every gene \(i\), a gene-level state is sampled:

\[
S_i \in \{00,10,01,11\}.
\]

The state probabilities are derived from the single-trait risk proportions
\(\pi_1^S\), \(\pi_2^S\), and the shared-risk proportion \(\pi_{11}\):

\[
\begin{aligned}
\pi_{00} &= 1-\pi_1^S-\pi_2^S+\pi_{11},\\
\pi_{10} &= \pi_1^S-\pi_{11},\\
\pi_{01} &= \pi_2^S-\pi_{11},\\
\pi_{11} &= \pi_{11}.
\end{aligned}
\]

As in the mTADA simulation, \(\pi_1^S=0.05\) and \(\pi_2^S=0.03\).

For an active trait \(t\) and annotation category \(c\), the gene-specific
relative risk is generated from:

\[
\gamma_{itc}\sim
\operatorname{Gamma}(\bar{\gamma}_{tc}\beta_{tc},\beta_{tc}).
\]

For an inactive trait, \(\gamma_{itc}=1\). Given mutation rate \(\mu_{ic}\),
trio count \(N_t\), and relative risk \(\gamma_{itc}\), the de novo count is:

\[
X_{itc}\sim
\operatorname{Poisson}(2N_t\mu_{ic}\gamma_{itc}).
\]

This is the mTADA de novo Poisson/Gamma mechanism. State \(11\) is assigned at
the gene level and does not require the same mutation category to drive both
traits.

## Analysis Workflow

For each simulated gene-category count, the analytic TADA de novo Bayes factor
is calculated:

\[
\mathrm{BF}_{itc}
=
\frac{
\operatorname{NB}\left(
X_{itc};
\bar{\gamma}_{tc}\beta_{tc},
\frac{\beta_{tc}}{\beta_{tc}+2N_t\mu_{ic}}
\right)
}{
\operatorname{Poisson}(X_{itc};2N_t\mu_{ic})
}.
\]

Category Bayes factors are combined within each gene on the log scale. The same
simulated counts and mutation rates are then analyzed by:

1. **Joint MIRAGE:** the proposed gene-level four-state model.
2. **Separate analyses:** independent single-trait posterior analyses.
3. **Original mTADA:** the public `DN2traits` Stan/VB implementation.

For this mTADA-based benchmark, category Bayes factors enter directly, which is
equivalent to fixing MIRAGE \(\eta=1\). This benchmark therefore tests the
joint four-state inference under the mTADA gene-category DGM. It does not
pretend that mTADA gene-level DNM counts are variant-level case-control data.

## Representative Run

One complete representative dataset was generated using the public mTADA
mutation rates and the following settings:

| Parameter | Value |
|---|---|
| Number of genes | 19,358 |
| Annotation categories | Damaging missense and LoF |
| Trait 1 risk proportion | 0.05 |
| Trait 2 risk proportion | 0.03 |
| Shared-risk proportion | 0.02 |
| Trait 1 trios | 2,000 |
| Trait 2 trios | 2,000 |
| Trait 1 mean RRs | 24 and 6 |
| Trait 2 mean RRs | 12 and 2 |
| Gamma dispersion parameter | 1 for each category |
| Simulation seed | 20200648 |
| Original mTADA VB iterations | 5,000 |

The true four-state proportions for this setting are:

\[
(\pi_{00},\pi_{10},\pi_{01},\pi_{11})
=(0.94,0.03,0.01,0.02).
\]

## Representative Results

### AUC

```{r representative-auc, echo=FALSE}
data.frame(
  method = c(
    "Joint MIRAGE", "Separate", "Original mTADA",
    "Joint MIRAGE", "Separate", "Original mTADA",
    "Joint MIRAGE", "Separate", "Original mTADA"
  ),
  target = rep(c("Trait 1", "Trait 2", "Pleiotropy"), each = 3),
  AUC = c(
    0.6687, 0.6587, 0.6685,
    0.6463, 0.5839, 0.6462,
    0.6886, 0.6887, 0.6887
  )
)
```

In this replicate, joint analysis improved Trait 2 AUC from 0.584 to 0.646.
Trait 1 AUC increased from 0.659 to 0.669. Joint MIRAGE and original mTADA
produced very similar rankings because both use the same four-state likelihood
structure in this benchmark.

### PP Threshold of 0.8

```{r representative-pp08, echo=FALSE}
data.frame(
  method = c(
    "Joint MIRAGE", "Separate", "Original mTADA",
    "Joint MIRAGE", "Separate", "Original mTADA",
    "Joint MIRAGE", "Separate", "Original mTADA"
  ),
  target = rep(c("Trait 1", "Trait 2", "Pleiotropy"), each = 3),
  discoveries = c(53, 44, 54, 9, 3, 13, 9, 0, 13),
  power = c(
    0.0514, 0.0444, 0.0524,
    0.0133, 0.0050, 0.0183,
    0.0193, 0, 0.0265
  ),
  empirical_FDP = c(
    0.0377, 0, 0.0370,
    0.1111, 0, 0.1538,
    0.1111, 0, 0.1538
  )
)
```

Joint MIRAGE found more Trait 1 and Trait 2 genes than the separate analyses.
For Trait 2, discoveries increased from 3 to 9. This gain came with one false
discovery in this replicate, so both power and empirical FDP must be evaluated
over repeated simulations.

### Mixing Proportion and Runtime

```{r representative-pi-runtime, echo=FALSE}
data.frame(
  method = c("Joint MIRAGE", "Original mTADA"),
  true_pi11 = c(0.02, 0.02),
  estimated_pi11 = c(0.0217, 0.0279),
  runtime_seconds = c(11.0, 108.9)
)
```

The separate analyses do not estimate a joint \(\pi_{11}\). The original mTADA
Stan VB fit completed, but its legacy `locfit` HPD routine raised
`HPD interval not closed` for this replicate. The reported mTADA estimate uses
a bounded density-mode fallback applied to the original `p12` VB draws. This
fallback is recorded in the task metadata rather than hidden.

Runtime excludes the one-time Stan compilation cost, which was approximately
85 seconds on the development computer.

### Bayesian FDR Check

At target Bayesian FDR 0.05:

```{r representative-bfdr, echo=FALSE}
data.frame(
  method = c(
    "Joint MIRAGE", "Separate", "Original mTADA",
    "Joint MIRAGE", "Separate", "Original mTADA"
  ),
  target = rep(c("Trait 1", "Trait 2"), each = 3),
  estimated_BFDR = c(
    0.0495, 0.0492, 0.0497,
    0.0358, 0.0439, 0.0410
  ),
  empirical_FDP = c(
    0.0217, 0, 0.0200,
    0.1250, 0, 0.1111
  )
)
```

This table is a diagnostic for one dataset, not an FDR-calibration result.
Calibration requires averaging empirical FDP across many replicates at each
simulation setting.

## Weak-Signal Sensitivity Run: mRR (5, 3) and (4, 2)

This additional run uses the same mTADA data-generating mechanism, public
gene-level mutation rates, four-state proportions, sample sizes, and random
seed as the representative analysis above. It does not replace that analysis.
Only the mean relative risks were reduced to create a deliberately weak-signal
setting:

| Parameter | Value |
|---|---|
| Genes | 19,358 |
| Trait 1 mean RRs (MiD, LoF) | (5, 3) |
| Trait 2 mean RRs (MiD, LoF) | (4, 2) |
| True pleiotropic proportion | 0.02 |
| Trait 1 and Trait 2 trios | 2,000 each |
| Simulation seed | 20200648 |
| Joint MIRAGE maximum EM iterations | 2,000 |
| EM tolerance | 1e-6 |

No model parameter was altered to force discoveries. Joint MIRAGE converged
naturally after 1,891 iterations. The observed-data log-likelihood was
non-decreasing, the final log-likelihood was 2.006309, and the final maximum
parameter change was 9.9973e-7. The estimated pleiotropic proportion was
0.003695 compared with the generating value of 0.02.

### Weak-Signal AUC

```{r weak-signal-auc, echo=FALSE}
data.frame(
  method = rep(c("Joint MIRAGE", "Separate", "Original mTADA"), 3),
  target = rep(c("Trait 1", "Trait 2", "Pleiotropy"), each = 3),
  AUC = c(
    0.5570, 0.5547, 0.5571,
    0.5697, 0.5410, 0.5667,
    0.5846, 0.5846, 0.5846
  )
)
```

The joint analysis modestly improved Trait 2 ranking relative to the separate
analysis in this replicate. The three methods had nearly identical
pleiotropic ranking AUC. These are single-replicate observations rather than
estimates of average performance.

### Posterior-Probability and BFDR Thresholds

```{r weak-signal-thresholds, echo=FALSE}
data.frame(
  method = c(
    "Joint MIRAGE", "Joint MIRAGE", "Joint MIRAGE",
    "Separate", "Separate", "Separate",
    "Original mTADA", "Original mTADA", "Original mTADA"
  ),
  target = rep(c("Trait 1", "Trait 2", "Pleiotropy"), 3),
  discoveries_at_PP_0.5 = c(0, 0, 0, 2, 0, 0, 2, 0, 0),
  discoveries_at_PP_0.7 = 0,
  discoveries_at_PP_0.8 = 0
)
```

At Bayesian FDR at most 0.05, all three methods also selected zero genes for
all three targets. Consequently, TPR and FPR are both zero at those fixed
thresholds. This does not indicate perfect classification: it indicates that
no posterior probability was high enough to form a discovery set under this
weak-signal configuration.

### Joint MIRAGE Top-K Ranking

When no genes cross a reporting threshold, top-K metrics still describe how
well true risk genes are enriched near the top of the ranking. TP@K and FP@K
are counts among the K highest posterior probabilities; TPR@K uses all true
risk genes as the denominator, while FPR@K uses all non-risk genes.

```{r weak-signal-top-k, echo=FALSE}
data.frame(
  target = rep(c("Trait 1", "Trait 2", "Pleiotropy"), each = 3),
  K = rep(c(50, 100, 200), 3),
  TP = c(10, 20, 33, 7, 12, 21, 9, 12, 18),
  FP = c(40, 80, 167, 43, 88, 179, 41, 88, 182),
  precision = c(0.200, 0.200, 0.165, 0.140, 0.120, 0.105,
                0.180, 0.120, 0.090),
  TPR = c(0.0101, 0.0202, 0.0333, 0.0116, 0.0200, 0.0349,
          0.0217, 0.0289, 0.0434),
  FPR = c(0.00218, 0.00436, 0.00909, 0.00229, 0.00469, 0.00954,
          0.00216, 0.00465, 0.00961)
)
```

For example, 9 of the top 50 pleiotropy-ranked genes were truly pleiotropic,
giving precision@50 of 0.18. These ranking results show weak enrichment, but
they do not turn the top-K lists into posterior-probability discoveries.

The simulation summary pipeline now records PP thresholds 0.5, 0.7, and 0.8;
BFDR diagnostics; and TP, FP, precision, TPR, and FPR at K = 50, 100, and 200
for every target and method. The prior result files remain unchanged.
## Full mTADA Paper Grid

The implemented full grid follows the parameter combinations shown in the
mTADA paper:

| Dimension | Values |
|---|---|
| Shared-risk proportion | \(\pi_{11}=0, 0.02\) |
| Trait 1 mean RRs | \((105,29)\) or \((24,6)\) |
| Trait 2 mean RRs | \((12,2)\) or \((87,23)\) |
| Trio counts | All nine pairs formed from 1,000, 2,000, and 5,000 |
| Repeated seeds | 100 per setting |

The full design contains:

\[
2\times4\times9\times100=7{,}200
\]

simulated datasets. For every dataset the pipeline records:

- AUC for Trait 1, Trait 2, and pleiotropy;
- power, TPR, FPR, precision, and empirical FDP at PP at least 0.5, 0.7, and 0.8;
- estimated BFDR and empirical FDP at targets 0.01, 0.05, 0.10, and 0.20;
- TP, FP, precision, TPR, and FPR at K = 50, 100, and 200;
- recovery of \(\pi_{11}\);
- runtime and fitting errors.

The runner uses one checkpoint per dataset and supports non-overlapping task
ranges for parallel execution. The full 7,200-dataset experiment has been
implemented but has not yet completed. Therefore, the representative result
above demonstrates that the data generation and all three analysis paths work;
it is not a final method comparison or calibration claim.

## Variant-Level MIRAGE-Style Exploration

After the mTADA-style gene-category simulation above, we also explored two
variant-level case-control simulations. These are closer to the original
single-trait MIRAGE simulation structure because the raw data contain one row
per variant rather than one row per gene-category count.

Both explorations used the same gene-level four-state truth:

| Quantity | Value |
|---|---:|
| Replicates | 30 |
| Genes per replicate | 500 |
| Variants per gene per trait | 15 |
| Variant rows per trait per replicate | 7,500 |
| Variant rows across two traits per replicate | 15,000 |
| Variant rows across 30 replicates | 450,000 |
| Exact gene-state counts per replicate | 470 `00`, 20 `10`, 5 `01`, 5 `11` |
| Trait 1 risk genes | 25 |
| Trait 2 risk genes | 10 |
| Pleiotropic genes | 5 |
| Cases and controls per trait | 3,000 cases and 3,000 controls |
| Annotation categories | `C1`, `C2`, `C3` |
| Category probabilities | 0.60, 0.30, 0.10 |
| Trait 1 eta values | 0.05, 0.20, 0.50 |
| Trait 2 eta values | 0.03, 0.15, 0.40 |
| Mean relative risks | 3, 3, 5 |
| Gamma dispersion | 1 for each category |

The methods compared were:

1. **Joint MIRAGE:** the two-trait MIRAGE model with \(\pi\) and \(\eta\)
   estimated.
2. **Separate MIRAGE:** two single-trait MIRAGE-style analyses with an
   independent four-state prior.
3. **Aggregated mTADA:** the original mTADA implementation applied after
   aggregating variant-level case counts and baseline rates to gene-category
   counts.

The aggregated mTADA analysis is included as a stress-test comparator. It is not
the native data format for this variant-level case-control design because it
does not use the control counts.

### Paper-Like Variant-Level Design

The first variant-level design follows the earlier MIRAGE-style setup:

- baseline allele rates were generated for each variant;
- active genes contained risk variants according to the category-specific
  \(\eta\) values;
- risk variants increased case counts through their relative risks;
- controls were generated from the same baseline rate without disease
  enrichment;
- there were no deliberately inflated noncausal background variants.

Mean AUC over 30 replicates:

```{r variant-paper-like-auc, echo=FALSE}
data.frame(
  method = c(
    "Joint MIRAGE", "Separate MIRAGE", "Aggregated mTADA",
    "Joint MIRAGE", "Separate MIRAGE", "Aggregated mTADA",
    "Joint MIRAGE", "Separate MIRAGE", "Aggregated mTADA"
  ),
  target = rep(c("Trait 1", "Trait 2", "Pleiotropy"), each = 3),
  mean_AUC = c(
    0.712, 0.704, 0.736,
    0.671, 0.629, 0.720,
    0.749, 0.732, 0.791
  )
)
```

Selected parameter learning:

```{r variant-paper-like-pi, echo=FALSE}
data.frame(
  method = c("Joint MIRAGE", "Separate MIRAGE", "Aggregated mTADA"),
  true_pi11 = c(0.010, 0.010, 0.010),
  mean_estimated_pi11 = c(0.048, 0.001, 0.011)
)
```

In this paper-like variant-level design, aggregated mTADA still had the highest
mean AUC. This is expected because after aggregation the main signal is still a
gene-category burden signal, which is favorable to mTADA.

### Control-Informative Variant-Level Design

The second variant-level design was constructed to represent a setting where
the control counts carry information that should not be discarded:

- the same 500 genes, 15 variants per gene, 30 replicates, and four-state truth
  were used;
- 18% of variants were assigned to a high-background noncausal group;
- those background variants had their case and control count-generating rate
  multiplied by 8;
- the mTADA aggregation still used the baseline variant rate, not the inflated
  count-generating rate;
- true disease variants increased case counts, while high-background variants
  increased both case and control counts.

This design creates variants that look large in case-count burden but are not
case-enriched relative to controls. MIRAGE can use the control counts to
down-weight them; aggregated mTADA cannot.

Mean AUC over 30 replicates:

```{r variant-control-informative-auc, echo=FALSE}
data.frame(
  method = c(
    "Joint MIRAGE", "Separate MIRAGE", "Aggregated mTADA",
    "Joint MIRAGE", "Separate MIRAGE", "Aggregated mTADA",
    "Joint MIRAGE", "Separate MIRAGE", "Aggregated mTADA"
  ),
  target = rep(c("Trait 1", "Trait 2", "Pleiotropy"), each = 3),
  mean_AUC = c(
    0.747, 0.741, 0.605,
    0.688, 0.666, 0.580,
    0.790, 0.778, 0.593
  )
)
```

At posterior probability at least 0.8:

```{r variant-control-informative-pp08, echo=FALSE}
data.frame(
  method = c(
    "Joint MIRAGE", "Separate MIRAGE", "Aggregated mTADA",
    "Joint MIRAGE", "Separate MIRAGE", "Aggregated mTADA",
    "Joint MIRAGE", "Separate MIRAGE", "Aggregated mTADA"
  ),
  target = rep(c("Trait 1", "Trait 2", "Pleiotropy"), each = 3),
  mean_discoveries = c(
    5.07, 4.50, 270.50,
    1.80, 1.23, 251.27,
    0.63, 0.03, 152.30
  ),
  empirical_FDP = c(
    0.084, 0.047, 0.939,
    0.128, 0.033, 0.976,
    0.106, 0.000, 0.988
  )
)
```

Selected parameter learning:

```{r variant-control-informative-pi, echo=FALSE}
data.frame(
  method = c("Joint MIRAGE", "Separate MIRAGE", "Aggregated mTADA"),
  true_pi11 = c(0.010, 0.010, 0.010),
  mean_estimated_pi11 = c(0.023, 0.001, 0.009)
)
```

Under this control-informative design, joint MIRAGE clearly outperformed the
aggregated mTADA comparator in ranking for both marginal trait discovery and
pleiotropy. Aggregated mTADA also produced many high-posterior discoveries with
very high empirical FDP because the high-background noncausal variants were
interpreted as disease burden after aggregation.

## Reproducible Commands

The full runner is `code/run_mtada_full_grid.R`. For example:

```{r full-grid-command, eval=FALSE}
Sys.setenv(
  MTADA_REFERENCE_DIR = "D:/path/to/mTADA",
  MIRAGE_TASK_START = "1",
  MIRAGE_TASK_END = "900",
  MIRAGE_RUN_ORIGINAL_MTADA = "true"
)
source("code/run_mtada_full_grid.R")
```

Completed checkpoints are summarized with:

```{r summarize-command, eval=FALSE}
source("code/summarize_mtada_full_grid.R")
```

## Internal Software Smoke Test

Before the mTADA-based benchmark was added, a small 120-gene synthetic test was
used only to verify EM convergence, posterior normalization, and end-to-end
execution. That internal test remains in the automated test suite, but it is
not the scientific simulation study reported on this page.

## Simulation Design Summary

| No. | Simulation | Data size | Core design | Main result | Interpretation |
|---:|---|---|---|---|---|
| 1 | mTADA-style gene-count | 19,358 genes, 2 categories | mTADA DNM Poisson/Gamma gene-category counts | Joint MIRAGE \(\approx\) mTADA \(>\) Separate | The joint layer is correct, but this design does not test MIRAGE's variant-level advantage. |
| 2 | Weak-signal mTADA-style | 19,358 genes, 100 reps | Mean RRs reduced to Trait 1 \((5,3)\), Trait 2 \((4,2)\) | All methods had low AUC and almost no discoveries | This is a difficult weak-signal setting rather than a MIRAGE-favorable setting. |
| 3 | Paper-like variant-level | 500 genes, 15 variants/gene/trait, 30 reps | Variant-level case/control data with ordinary burden-like signal | mTADA \(>\) Joint MIRAGE \(>\) Separate | Aggregation preserves most of the signal, so mTADA remains strong. |
| 4 | Control-informative variant-level | 500 genes, 15 variants/gene/trait, 30 reps | 18% high-background noncausal variants added | Joint MIRAGE \(\gg\) mTADA | Control counts become informative; MIRAGE can down-weight background variants and gains a clear advantage. |

## Reference

Nguyen TH, Dobbyn A, Brown RC, et al. mTADA is a framework for identifying risk
genes from de novo mutations in multiple traits. *Nature Communications*.
2020;11:2929. doi:10.1038/s41467-020-16487-z.
